---
title: "Analysis of Optimal Treatment Rates"
output: html_document
date: "2026-10-08"
---

```{r setup, include=FALSE}
install_if_missing <- function(pkgs){
  to_install <- pkgs[!(pkgs %in% installed.packages()[, "Package"])]
  if(length(to_install)) install.packages(to_install)
}
install_if_missing(c("ordinal", "emmeans", "broom.mixed", "DHARMa", "dplyr", "ggplot2", "nnet", "tidyverse", "rms", "PermAlgo", "knitr", "mediation"))

knitr::opts_chunk$set(echo = TRUE, warning = FALSE)

library(emmeans)     # predicted probabilities and contrasts
library(ggplot2)
library(lme4)
library(tidyverse)
if (!("coxed" %in% installed.packages()[, "Package"])) install.packages("./coxed",repos=NULL, type="source")
library(coxed)

set.seed(1)

analysis_date <- "261008"
```

## Loading Data

```{r load_data, include=FALSE}
decisions <- read_csv(paste0("data/decisions_with_relative_quality_", analysis_date, ".csv"))%>% 
  mutate(Decision = decision_num,
         Case = as.factor(Case),
         ResponseId = as.factor(ResponseId)) %>%
  dplyr::select(ResponseId, AIInterface, Case, Decision, QualityScore, MinQuality, MaxQuality, RelativeQuality, IsBestScore)
responses <- read_csv(paste0("data/cleaned_respondent_level_", analysis_date, ".csv"))
quality <- read_csv("data/all_quality_ratings.csv") %>% 
  mutate(QualityScore = Final) %>%
  dplyr::select(-E1, -E2, -E3, -E4, -E5, -Final)
quality_scores <- read_csv("data/all_quality_ratings.csv") %>% 
  rename(QualityScore = Final) %>%
  dplyr::select(Case, Decision, QualityScore)
ai_concordance_types <- read_csv("data/AI_concordance_types.csv") %>%
  rename(Decision = decision_num) %>%
  left_join(quality_scores, by = c("Case", "Decision"))
```

```{r max_quality}
max_quality <- quality %>%
  group_by(Case) %>%
  summarise(MaxQuality = max(QualityScore), MinQuality = min(QualityScore), .groups = "drop")

# Add max score to quality table
quality <- quality %>%
  left_join(max_quality, by = "Case")
```


```{r}
smoothing_factor <- 0.05

cases <- unique(quality$Case)
interfaces <- unique(decisions$AIInterface)

decision_dists <- list()

for (case in cases) {

  # empirical distribution from No AI
  empirical <- decisions %>%
    filter(AIInterface == "No AI", Case == case) %>%
    count(Decision) %>%
    mutate(prob = n / sum(n)) %>%
    dplyr::select(Decision, prob)

  possible_decs <- quality %>%
    filter(Case == case) %>%
    distinct(Decision)

  smoothing <- possible_decs %>%
    mutate(prob = 1 / n())

  base_dist <- full_join(
    smoothing %>% mutate(prob = prob * smoothing_factor),
    empirical %>% mutate(prob = prob * (1 - smoothing_factor)),
    by = "Decision"
  ) %>%
    replace_na(list(prob.x = 0, prob.y = 0)) %>%
    mutate(prob = prob.x + prob.y) %>%
    dplyr::select(Decision, prob)

  for (iface in interfaces) {

    dist <- base_dist
    
    if (iface == "No AI") {
      dist <- dist %>% mutate(prob = prob / sum(prob))
      decision_dists[[paste(case, iface, sep = "|")]] <- dist
      next
    }

    matches <- ai_concordance_types %>%
      filter(AIInterface == iface, Case == case)
    mentioned <- matches %>% filter(ConcordanceType == "Mentioned")
    adjacent <- matches %>% filter(ConcordanceType == "Adjacent")
    other <- matches %>% filter(ConcordanceType == "None")
    non_mentioned <- matches %>% filter(ConcordanceType != "Mentioned")

    ### Aligned
    
    if (iface == "Options to Avoid") {
      aligned_dist <- dist %>% filter(Decision %in% non_mentioned$Decision)
    } else {
      aligned_dist <- dist %>% filter(Decision %in% mentioned$Decision)
    }

    decision_dists[[paste(case, iface, "Aligned", sep = "|")]] <- aligned_dist %>%
      mutate(prob = prob / sum(prob))
    
    ### Mentioned
    
    mentioned_dist <- dist %>% filter(Decision %in% mentioned$Decision)
    decision_dists[[paste(case, iface, "Mentioned", sep = "|")]] <- mentioned_dist %>%
      mutate(prob = prob / sum(prob))
    
    ### Adjacent
    
    adjacent_dist <- dist %>% filter(Decision %in% adjacent$Decision)
    decision_dists[[paste(case, iface, "Adjacent", sep = "|")]] <- adjacent_dist %>%
      mutate(prob = prob / sum(prob))
    
    ### Other
    
    other_dist <- dist %>% filter(Decision %in% other$Decision)
    decision_dists[[paste(case, iface, "None", sep = "|")]] <- other_dist %>%
      mutate(prob = prob / sum(prob))
  }
}
```

```{r sample_decision}
sample_decision <- function(case, iface, concordance = "Aligned", uniform = FALSE) {

  if (iface == "No AI") {
    dist <- decision_dists[[paste(case, iface, sep="|")]]
  } else {
    dist <- decision_dists[[paste(case, iface, concordance, sep="|")]]
  }
  
  if (nrow(dist) == 1) {
    return(as.character(dist$Decision))
  }

  if (uniform) {
    return(as.character(sample(dist$Decision, 1)))
  }
  as.character(sample(dist$Decision, 1, prob = dist$prob))
}
```

## Human-Only Model


```{r human_data_prep}

human_df <- decisions %>%
  mutate(Case = factor(Case),
         ResponseId = factor(ResponseId),
         AIInterface = relevel(factor(AIInterface), ref = "No AI"))

```

```{r human_model}

model <- glmer(
  IsBestScore ~ AIInterface + (1 | Case) + (1 | ResponseId),
  data = human_df,
  family = binomial
)

print(summary(model))

print("Test that all three weight coefficients are equal:")
coef_names <- c("AIInterfaceRecommended Treatment", "AIInterfaceOptions to Consider", "AIInterfaceOptions to Avoid")

# extract coefficient vector and covariance matrix for those terms
beta_all <- fixef(model)
vcov_all <- vcov(model)

L <- rbind(
  c(1, -1, 0),
  c(1,  0, -1)
)
colnames(L) <- coef_names

beta <- beta_all[coef_names]
V <- vcov_all[coef_names, coef_names, drop = FALSE]

# basic sanity checks
if (any(is.na(beta)) || any(is.na(V))) stop("One or more names in coef_names not found in model. Check names(coef(model_cond_only)).")
if (nrow(V) != length(beta)) stop("Covariance submatrix has wrong dimensions.")

Lb <- as.numeric(L %*% beta)
LVL <- L %*% V %*% t(L)

W <- as.numeric(t(Lb) %*% solve(LVL) %*% Lb)
df <- nrow(L)
p_value <- 1 - pchisq(W, df)

cat("Wald chi-square =", format(W, digits = 6), "on", df, "df; p =", format(p_value, digits = 6), "\n")

print("Emmeans:")

emm <- emmeans(model, ~ AIInterface)

ctr <- emmeans::contrast(emm, method = "revpairwise", adjust = "tukey")
ctr_df <- as.data.frame(ctr)
ctr_df$OR <- exp(ctr_df$estimate)
ctr_df$lower.OR <- exp(ctr_df$estimate - 1.96 * ctr_df$SE)
ctr_df$upper.OR <- exp(ctr_df$estimate + 1.96 * ctr_df$SE)

# tidy output
ctr_df <- ctr_df[, c("contrast", "estimate", "SE", "OR", "lower.OR", "upper.OR", "z.ratio", "p.value")]
print(ctr_df)

write.csv(ctr_df, "result_data/human_best_choice_ors_fixed_threshold.csv")
```

Run a model controlling for the expected quality of following the AI:
```{r control_expected_qual}
model <- glmer(
  IsBestScore ~ AIInterface + RelativeQuality + (1 | Case) + (1 | ResponseId),
  data = human_df,
  family = binomial
)

print(summary(model))

print("Emmeans:")

emm <- emmeans(model, ~ AIInterface)

ctr <- emmeans::contrast(emm, method = "revpairwise", adjust = "tukey")
ctr_df <- as.data.frame(ctr)
ctr_df$OR <- exp(ctr_df$estimate)
ctr_df$lower.OR <- exp(ctr_df$estimate - 1.96 * ctr_df$SE)
ctr_df$upper.OR <- exp(ctr_df$estimate + 1.96 * ctr_df$SE)

# tidy output
ctr_df <- ctr_df[, c("contrast", "estimate", "SE", "OR", "lower.OR", "upper.OR", "z.ratio", "p.value")]
print(ctr_df)

write.csv(ctr_df, "result_data/human_best_choice_ors_quality_control_fixed_threshold.csv")
```

## Run Bootstrap

We'll use the bootstrap to estimate the theoretical decision quality of always following the AI and provide uncertainty estimates. Specifically we want to estimate (a) the rate of choosing the highest-scoring decision, (b) the confidence intervals around these empirical rates, and (c) whether humans perform differently than the theoretical performance.

```{r bootstrap_setup}
response_ids <- unique(decisions$ResponseId)
cases <- unique(decisions$Case)
interfaces <- unique(decisions$AIInterface)
```

```{r bootstrap, include=FALSE, warning=FALSE}
bootstrapper <- function (concordance = "Aligned", B = 1000, uniform = FALSE, quality_cases = NULL) {
  means_results <- list()
  comparison_results <- list()

  for (b in 1:B) {
    if (b %% 100 == 0) {
      print(b)
    }
    
    # Resample clinicians
    boot_ids <- sample(response_ids, length(response_ids), replace = TRUE)
  
    # -----------------------------
    # Simulate AI-following behavior
    # -----------------------------
  
    boot_responses <- tibble(ResponseId = boot_ids) %>%
      left_join(decisions %>% dplyr::select(ResponseId, AIInterface, Case, RelativeQuality), by = "ResponseId", relationship = "many-to-many") %>%
      mutate(ResponseId = paste0(ResponseId, "_boot", row_number())) %>%
      dplyr::select(ResponseId, AIInterface, Case, RelativeQuality)
    
    sim_data <- boot_responses
    
    if (!is.null(quality_cases) && quality_cases == "high") {
      sim_data <- sim_data %>% filter(RelativeQuality >= 0)
    } else if (!is.null(quality_cases) && quality_cases == "low") {
      sim_data <- sim_data %>% filter(RelativeQuality < 0)
    }

    sim_data <- sim_data %>%
      filter(AIInterface != "No AI") %>%
      mutate(
        Decision = map2_chr(
          Case,
          AIInterface,
          ~ sample_decision(.x, .y, concordance = concordance, uniform = uniform)
        ),
        HumanAI = "Simulated"
      )
  
    sim_data <- sim_data %>%
      left_join(quality %>% mutate(Decision = as.character(Decision)), by = c("Case", "Decision")) %>%
      mutate(IsBestScore = QualityScore == MaxQuality)
  
    # -----------------------------
    # Prepare real decisions
    # -----------------------------
  
    if (concordance == "Aligned") {
      boot_real <- map_dfr(seq_along(boot_ids), function(i) {
        rid <- boot_ids[i]
      
        decisions %>%
          filter(ResponseId == rid) %>%
          mutate(
            ResponseId = paste0(rid, "_boot", i),  # ensure unique random effect level
            HumanAI = "Human"
          )
      
      })
    
      boot_real <- boot_real %>%
        mutate(Decision = as.character(Decision))

      # -----------------------------
      # Combine real + simulated
      # -----------------------------
    
      combined <- bind_rows(boot_real, sim_data)
    
    } else {
      combined <- sim_data
    }
    
    # -----------------------------
    # Empirical means (for computing CIs)
    # -----------------------------
  
    means_results[[b]] <- combined %>%
      group_by(AIInterface, HumanAI) %>%
      summarise(prob = mean(IsBestScore == TRUE),
                .groups = "drop")
    
    # if (compare_interfaces) {
    #   if (concordance == "Aligned") stop("Does not support comparing interfaces for Aligned")
    #   
    #   # Fit a logistic regression model
    #   combined <- combined %>% 
    #     mutate(AIInterface = relevel(as.factor(AIInterface), ref = "Recommended Treatment"),
    #            IsBestScore = as.factor(IsBestScore))
    #   model <- glmer(IsBestScore ~ AIInterface + (1 | ResponseId) + (1 | Case),
    #                  data = combined,
    #                  family = binomial)
    #   
    #   emm <- emmeans(model, ~ AIInterface, type = "response")
    #   comparison_results[[b]] <- as.data.frame(pairs(emm, adjust = "none", combine = TRUE))
    # }
  }

  bind_rows(means_results, .id = "bootstrap")  
  # if (compare_interfaces) {
  #   list("means" = bind_rows(means_results, .id = "bootstrap"),
  #        "comparisons" = bind_rows(comparison_results, .id = "bootstrap"))
  # } else {
  #   list("means" = bind_rows(means_results, .id = "bootstrap"),
  #        "comparisons" = NULL)
  # }
}
```


Use bootstrapping to get the means and CIs for alignment with AI:

```{r aligned_bootstrap}
set.seed(1)
bootstrap_df <- bootstrapper(concordance = "Aligned", B = 1000)

mean_summary <- bootstrap_df %>%
  filter(!is.na(prob)) %>%
  group_by(AIInterface, HumanAI) %>%
  summarise(
    mean = mean(prob),
    # Use bias-corrected confidence intervals
    ci.lower = bca(prob)[1],
    ci.upper = bca(prob)[2],
    .groups = "drop"
  )

write.csv(mean_summary, "result_data/best_choice_bootstrap_mean_cis.csv")
mean_summary
```

```{r}
write.csv(mean_summary, "result_data/best_choice_bootstrap_mean_cis.csv")
mean_summary
```
# Agreement

```{r agreement}
print("\n\n####### Analysis of concordance ########\n\n")

AI_labels <- c("Recommended Treatment", "Options to Consider", "Options to Avoid")

concordance_ors <- list()
concordance_dids <- list()
concordance_quality_ors <- list()

for (AI_label in AI_labels) {
  
  concordance_df <- read.csv(paste("data/advice_quality_", AI_label, "_260409.csv", sep = ""), stringsAsFactors = TRUE)

  concordance_df <- concordance_df %>%
    mutate(
      HasAI = as.factor(AIInterface != "No AI"),
      AdjacentConcordance = as.factor(ConcordanceType == "Adjacent"),
      MentionConcordance = as.factor(ConcordanceType == "Mentioned"),
      NoConcordance = as.factor(ConcordanceType == "None"),
      ConcordanceType = relevel(as.factor(ConcordanceType), ref = "None"),
    ) %>%
    dplyr::select(
      ResponseId, Case, HasAI, SimulatedAIHighQuality,
      MentionConcordance, AdjacentConcordance, NoConcordance, ConcordanceType, IsBestScore
    )
  
  mention_model <- glmer(MentionConcordance ~ HasAI * SimulatedAIHighQuality + (1 | Case) + (1 | ResponseId),
                      data = concordance_df,
                      family = "binomial")
  mention_base <- glmer(MentionConcordance ~ SimulatedAIHighQuality + (1 | Case) + (1 | ResponseId),
                         data = concordance_df,
                         family = "binomial")
  adj_model <- glmer(AdjacentConcordance ~ HasAI * SimulatedAIHighQuality + (1 | Case) + (1 | ResponseId),
                         data = concordance_df,
                         family = "binomial")
  adj_base <- glmer(AdjacentConcordance ~ SimulatedAIHighQuality + (1 | Case) + (1 | ResponseId),
                        data = concordance_df,
                        family = "binomial")
  none_model <- glmer(NoConcordance ~ HasAI * SimulatedAIHighQuality + (1 | Case) + (1 | ResponseId),
                     data = concordance_df,
                     family = "binomial")
  none_base <- glmer(NoConcordance ~ SimulatedAIHighQuality + (1 | Case) + (1 | ResponseId),
                    data = concordance_df,
                    family = "binomial")
  mention_marg <- glmer(MentionConcordance ~ HasAI + (1 | Case) + (1 | ResponseId),
                         data = concordance_df,
                         family = "binomial")
  adj_marg <- glmer(AdjacentConcordance ~ HasAI + (1 | Case) + (1 | ResponseId),
                        data = concordance_df,
                        family = "binomial")
  none_marg <- glmer(NoConcordance ~ HasAI + (1 | Case) + (1 | ResponseId),
                    data = concordance_df,
                    family = "binomial")
  
  print(anova(mention_model, mention_base))
  print(anova(adj_model, adj_base))
  print(anova(none_model, none_base))
  ors <- bind_rows(
    as.data.frame(
      pairs(emmeans(mention_model, ~ HasAI | SimulatedAIHighQuality, mode = "latent"), 
            simple = "HasAI", reverse = TRUE, adjust = "none")) 
    %>% 
      mutate(ConcordanceType = "Mentioned"), 
    as.data.frame(
      pairs(emmeans(mention_marg, ~ HasAI, mode = "latent"), 
            simple = "HasAI", reverse = TRUE, adjust = "none")) 
    %>% 
      mutate(ConcordanceType = "Mentioned", All = TRUE), 
    as.data.frame(
      pairs(emmeans(adj_model, ~ HasAI | SimulatedAIHighQuality, mode = "latent"), 
            simple = "HasAI", reverse = TRUE, adjust = "none")) 
    %>% 
      mutate(ConcordanceType = "Adjacent"),
    as.data.frame(
      pairs(emmeans(adj_marg, ~ HasAI, mode = "latent"), 
            simple = "HasAI", reverse = TRUE, adjust = "none")) 
    %>% 
      mutate(ConcordanceType = "Adjacent", All = TRUE)
    # as.data.frame(
    #   pairs(emmeans(none_model, ~ HasAI | SimulatedAIHighQuality, mode = "latent"), 
    #         simple = "HasAI", reverse = TRUE, adjust = "none")) 
    # %>% 
    #   mutate(ConcordanceType = "None"),
    # as.data.frame(
    #   pairs(emmeans(none_marg, ~ HasAI, mode = "latent"), 
    #         simple = "HasAI", reverse = TRUE, adjust = "none")) 
    # %>% 
    #   mutate(ConcordanceType = "None", All = TRUE)
    )
  est <- ors[, "estimate"]
  se  <- ors[, "SE"]
  ors$odds_ratio  <- exp(est)
  z <- qnorm(0.975)
  ors$ci_low <- exp(est - z * se)
  ors$ci_high <- exp(est + z * se)
  concordance_ors[[AI_label]] = ors %>% mutate(AIInterface = AI_label)
  
  dids <- rbind(
    as.data.frame(
      emmeans::contrast(emmeans(mention_model, ~ HasAI * SimulatedAIHighQuality), mode = "latent", interaction = "consec", adjust = "none"))
    %>% 
      mutate(ConcordanceType = "Mentioned"), 
    as.data.frame(
      emmeans::contrast(emmeans(adj_model, ~ HasAI * SimulatedAIHighQuality), mode = "latent", interaction = "consec", adjust = "none"))
    %>% 
      mutate(ConcordanceType = "Adjacent")
    # as.data.frame(
    #   emmeans::contrast(emmeans(none_model, ~ HasAI * SimulatedAIHighQuality), mode = "latent", interaction = "consec", adjust = "none")) 
    # %>% 
    #   mutate(ConcordanceType = "None")
  )
  est <- dids[, "estimate"]
  se  <- dids[, "SE"]
  dids$odds_ratio  <- exp(est)
  z <- qnorm(0.975)
  dids$ci_low <- exp(est - z * se)
  dids$ci_high <- exp(est + z * se)
  
  concordance_dids[[AI_label]] <- dids %>% mutate(AIInterface = AI_label)
  
  AI_only_df <- concordance_df %>% filter(HasAI == TRUE & ConcordanceType != "Mentioned")
  quality_model <- glmer(IsBestScore ~ ConcordanceType + (1 | Case) + (1 | ResponseId),
                        data = AI_only_df,
                        family = binomial)
  
  quality_model_simple <- glmer(IsBestScore ~ (1 | Case) + (1 | ResponseId),
                               data = AI_only_df,
                               family = binomial)
  quality_model_bigger <- glmer(IsBestScore ~ ConcordanceType * SimulatedAIHighQuality + (1 | Case) + (1 | ResponseId),
                               data = AI_only_df,
                               family = binomial)
  
  print(anova(quality_model, quality_model_bigger))

  emm <- emmeans(quality_model_bigger, ~ ConcordanceType | SimulatedAIHighQuality, type = "latent")
  ctr <- as.data.frame(pairs(emm, simple = "ConcordanceType", combine = TRUE, reverse = TRUE, adjust = "none"))
  est <- ctr[, "estimate"]
  se  <- ctr[, "SE"]
  ctr$odds_ratio  <- exp(est)
  z <- qnorm(0.975)
  ctr$ci_low <- exp(est - z * se)
  ctr$ci_high <- exp(est + z * se)
  concordance_quality_ors[[AI_label]] <- as.data.frame(ctr) %>% mutate(AIInterface = AI_label)
}
concordance_ors_combined <- bind_rows(concordance_ors)
concordance_ors_combined$p.value.adj <- p.adjust(concordance_ors_combined$p.value, method = "holm")
concordance_ors_combined <- concordance_ors_combined %>%
  dplyr::select(AIInterface, ConcordanceType, contrast, SimulatedAIHighQuality, odds_ratio, ci_low, ci_high, p.value.adj)
print(concordance_ors_combined)
concordance_dids_combined <- bind_rows(concordance_dids)
concordance_dids_combined$p.value.adj <- p.adjust(concordance_dids_combined$p.value, method = "holm")
concordance_dids_combined <- concordance_dids_combined %>% 
  dplyr::select(AIInterface, ConcordanceType, HasAI_consec, SimulatedAIHighQuality_consec, odds_ratio, ci_low, ci_high, p.value.adj)
print(concordance_dids_combined)
concordance_quality_ors_combined <- bind_rows(concordance_quality_ors)
concordance_quality_ors_combined$p.value.adj <- p.adjust(concordance_quality_ors_combined$p.value, method = "holm")
concordance_quality_ors_combined <- concordance_quality_ors_combined %>% 
  dplyr::select(AIInterface, SimulatedAIHighQuality, contrast, odds_ratio, ci_low, ci_high, p.value.adj)
print(concordance_quality_ors_combined)
```

```{r write_agreement_outputs}
write.csv(concordance_ors_combined, "result_data/concordance_ors.csv")
write.csv(concordance_dids_combined, "result_data/concordance_dids.csv")
write.csv(concordance_quality_ors_combined, "result_data/concordance_quality_ors.csv")
```
